---
title: "CoxPropHazard"
format: html
editor: visual
---

### Example Model

```{r}
library(survival)
data <- survival::colon
head(data)
```

**Colon dataset summary**

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/colon.html

|           |          |                                                                                   |
|----------------|----------------|----------------------------------------|
| Field     | DataType | Description                                                                       |
| id:       | num      | id                                                                                |
| study:    | num      | 1 for all patients                                                                |
| rx:       | factor   | Treatment - Obs(ervation), Lev(amisole), Lev(amisole)+5-FU                        |
| sex:      | num      | 1=male                                                                            |
| age:      | num      | in years                                                                          |
| obstruct: | num      | obstruction of colon by tumour                                                    |
| perfor:   | num      | perforation of colon                                                              |
| adhere:   | num      | adherence to nearby organs                                                        |
| nodes:    | num      | number of lymph nodes with detectable cancer                                      |
| time:     | num      | days until event or censoring                                                     |
| status:   | num      | censoring status                                                                  |
| differ:   | num      | differentiation of tumour (1=well, 2=moderate, 3=poor)                            |
| extent:   | num      | Extent of local spread (1=submucosa, 2=muscle, 3=serosa, 4=contiguous structures) |
| surg:     | num      | time from surgery to registration (0=short, 1=long)                               |
| node4:    | num      | more than 4 positive lymph nodes                                                  |
| etype:    | num      | event type: 1=recurrence,2=death                                                  |

## Assumptions of Cox Proportional Hazards

#### Proportional Hazards Assumption

Hazard ratios should be constant over time.

#### Linearity

relationship between covariates and the log hazard is linear.

#### Independence of Survival Times

Each data point would need to be independent of each other.

#### No Omitted Covariates

All need to be included or the model will be biased

#### Model Creation

Surv(time, event) - Creates survival object with time and event

variables after \~ are covariates

```{r}
cox <- coxph(Surv(time, status) ~ age + sex + node4,
             data = data)

summary(cox)
```

### Testing Assumptions

#### Proportional Hazards Assumption

```{r}
cox.zph(cox)

plot(cox.zph(cox))
```

Interpretation of output

-   P value \<= .05 indicates the assumption is violated for that covariate.

-   The visuals display Shoenfeld residuals over time for each covariate. If the residuals are in a straight line, the assumption is satisfied.

-   If the line changes over time it may indicate the hazard ratio changes over time for the covariate

-   No clear pattern in the plot indicates no pattern in the residuals. The plot is inconclusive and refer to the p value

If the assumption is broken:

-   Stratification - analyzing the data separately for different covariates that violate the assumption. Use strata() within coxph() when making the model

-   Transformation - a transformation, such as log, can be applied to a covariate that breaks the assumption

#### Linearity Test

Tested by plotting Martingale Residuals against continuous covariates.

```{r}
martingale <- resid(cox, type = "martingale")

#Create a plot for each covariate in the model
par(mfrow = c(1,3))
plot(data$age, martingale, main = "Martingale Residuals: age")
abline(h = 0, col = "red")
plot(data$sex, martingale, main = "Martingale Residuals: sex")
abline(h = 0, col = "red")
plot(data$nodes, martingale, main = "Martingale Residuals: nodes")
abline(h = 0, col = "red")

```

-   Increasing or decreasing slope indicates a violation of the linearity assumption

-   A horizontal band around 0 indicates the assumption is not violated

-   Points close to zero indicate the model fits the observations well and vice versa for points far from zero

-   More points above zero indicates the model is underestimating the hazard for a significant number of observations.

    If this assumption is violated:

-   For the covariate in question, square the term in the model

-   Use splines for the covariate in question. From the splines library, use ns(covariate, df = \_ )

-   If the variable is continuous, convert it into a categorical variable.

### Evaluation

#### C_index

Concordant if the individual with a higher predicted risk experienced the event before the individual with the lower expected risk.

Discordant if the individual with the higher predicted risk experiences the event after the individual with the lower predicted risk.

Tied if the individuals experienced the event at the same time or did not experience the event in the period.

The c index is between 0 and 1.

1 means the model perfectly discriminates between individuals

0 means the model is completely unable to discriminate between individuals.

.5 means the model performs at random chance.

Interpretation of c index. A c index of \_\_\_ means that the model correctly predicts the relative risk of \_\_\_ of the comparable pairs.

$$
C-index = \frac{Number of Concordant Pairs}{Pairs}
$$

```{r}
#Get C index
(c_index <- concordance(cox, newdata = data))
```

### Visualizations

Survival curves are a common way to represent Cox models.

#### Using the ggsurvplot Function to Create Survival Curve

```{r}
library(survminer)
library(tidyverse)

survfit <- survfit(cox)

ggsurvplot(survfit,
           data = data,
           conf.int = TRUE,
           palette = "Dark1",
           pval = TRUE)
```

#### Using ggplot to Create Hazard Ratio 95% CI

```{r}
library(broom)

model_summary <- tidy(cox)

ggplot(model_summary,
       aes(x = term, y = exp(estimate), 
           ymin = exp(estimate-1.96*std.error),
           ymax = exp(estimate+1.96*std.error)))+
  geom_point()+
  geom_errorbar(width = .2)+
  scale_y_continuous(trans = "log")+
  labs(x = "Covariates",
       y = "Hazard Ratio (95% CI)",
       title = "Hazard Ratios from Model")+theme_bw()
  
```

### Predictions

```{r}
#Create training and testing sets

set.seed(100)

split <- sample(1:nrow(data), size = round(.9*nrow(data)))

train <- data[split,]
test <- data[-split,]
```

**Create Model for Training Data Set**

```{r}
m <- coxph(Surv(time, status) ~ strata(age) + sex + strata(obstruct) + perfor + surg,
             data = train)

summary(m)
```

#### **Check Assumptions**

**Proportional Hazards**

```{r}
cox.zph(m)

plot(cox.zph(m))
```

```{r}
m_martingale <- resid(m, type = "martingale")

#Create a plot for each covariate in the model
par(mfrow = c(1,5))
plot(train$age, m_martingale, main = "Martingale Residuals: age")
abline(h = 0, col = "red")

plot(train$sex, m_martingale, main = "Martingale Residuals: sex")
abline(h = 0, col = "red")

plot(train$obstruct, m_martingale, main = "Martingale Residuals: nodes")
abline(h = 0, col = "red")

plot(train$perfor, m_martingale, main = "Martingale Residuals: nodes")
abline(h = 0, col = "red")

plot(train$surg, m_martingale, main = "Martingale Residuals: nodes")
abline(h = 0, col = "red")
```

```{r}
(m_c_index <- concordance(cox, newdata = test))

```

```{r}
test$loghazard <- predict(m, newdata = test,type = "risk")

test$hazardratios <- exp(test$loghazard)

head(select(test,loghazard,hazardratios,age,sex,obstruct,perfor,surg))
```

### 
